---
title: "Cigarette consumption"
author: "Duncan Gillespie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cigarette consumption}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
bibliography: consumption-refs.bib
header-includes:
  \usepackage{float}
  \usepackage{amsmath}
link-citations: yes
citation_package: natbib
biblio-style: vancouver
urlcolor: blue
---

```{r setup, include = FALSE, results = 'hide', warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.pos = 'H'
)

#library(readxl)
#library(data.table)
suppressPackageStartupMessages(library(ggplot2))
#library(cowplot)

#knitr::opts_knit$set(root.dir = "/Volumes/Shared/ScHARR/PR_STAPM/Code/R_packages/tobalcepi")
#knitr::opts_knit$set(root.dir = "X:/ScHARR/PR_STAPM/Code/R_packages/tobalcepi")

```

# Introduction
The Health Survey for England (HSE) is a series of annual surveys covering health and health-related behaviours. For the Sheffield Tobacco Policy Model (STPM) we use data from years 2001 to the latest available. We use these data to inform the trends in smoking prevalence, the socio-demographic variation in smoking prevalence, and as inputs to a procedure that we use to infer the age-specific probabilities of smoking initiation and quitting (see the `smoke.trans` R package). Our upper age limit is 89 years, but otherwise we make use of all ages and incorporate all booster samples (e.g. in 2004 there was an ethnicity booster sample).  

To support the processing of the HSE data, there is the `hseclean` R package, which contains a collection of functions to read, process and summarise the HSE data. The purpose of this vignette is to explain how we use the HSE data to inform the patterns of cigarette consumption, and to explain how the `hseclean` package supports this.  

# Demographic and socio-economic variables
The `hseclean` package contains a range of functions to clean covariates in the data, which are explained more thoroughly in the vignette on covariates. Here we describe the important things to consider for the processing of smoking data.  

# Cigarette smoking variables
Questions about cigarette smoking have been asked of adults aged 16 and over as part of the HSE series since 1991. We use data from 2001 to the latest year available. We use data on children (12-15 years) and adults (16+ years). There is often a special section in the annual HSE report devoted to describing trends in cigarette smoking e.g. \href{https://files.digital.nhs.uk/publicationimport/pub22xxx/pub22610/hse2015-adult-smo.pdf}{HSE 2015}.  

## Cigarette smoking status
The function `smk_status()` categorises cigarette smoking into current, former and never regular cigarette smokers. If some smokes either regularly or ocassionally, then they are classified as a current regular cigarette smoker. People who used to smoke regularly or ocassionally are classified as former smokers, but people who have only tried a cigarette once or twice are classified as never smokers. We create this variable for children aged 8-15 years and adults aged >= 16 years. Ever-smokers are people who are either current or former smokers.  

## Former smoking
The function `smk_former()` cleans the data on the time since quitting and time spent as a regular smoker among former smokers. The main issue to be overcome is that in the HSE 2015+, time since quit and time spent as a smoker is provided in categories rather than single years. We simulate the single years by just picking a value at random within the time interval, using `num_sim()`. We also fill missing data: 

- For children 8-15 years, we assume that missing values for former smokers = 1 year.
- For adults, we fill missing values with the average value for each age, sex and IMD quintile subgroup.  

## Smoking life-histories
The function `smk_life_history()` cleans the ages that define when smokers started and stopped being regular cigarette smokers. For each individual smoker, the data recorded in the HSE implies a single age at which a smoker started to smoke and, if they stopped, an age at which they did so. This provides a simplified view of what might be a complicated life history of smoking, e.g. smoking to different frequencies or levels, or starting and stopping multiple times.  

Both the start age and stop age will have error in them e.g. due to uncertainty in respondent recall, and, for years 2015+, due to the reporting in categories of time intervals rather than single years, which we then impute introducing random error. Start age is likely to be biased towards earlier ages, because for adults with missing values we use the age first tried a cigarette, and for children the variable for start age does not necessarily mean the start of regular smoking, it is just the age at which they started to smoke.

We also create a variable for the age at which an individual was censored from our data sample - this is their age at the survey + 1 year.  

Any missing data is assigned the average start or stop age for each age, sex and IMD quintile.  

## Amount and type of cigarette smoked by current smokers
The function `smk_amount()` cleans the variables that describe how much, what and to what level of addiction people smoke. The main variable is the average number of cigarettes smoked per day. For adults this is calculated from questions about how many cigarettes are smoked typically on a weekday vs. a weekend. For children, this is based on asking how many cigarettes were smoked in the last week. Missing values are imputed as the average amount smoked for an age, sex and IMD quintile subgroup.  

We categorise cigarette preferences based on the answer to 'what is the main type of cigarette smoked'. In later years of the Health Survey for England, new questions are added that ask how many handrolled vs. machine rolled cigarettes are smoked on a weekday vs. a weekend. We currently don't use those questions because they were not asked in all years.  

We also categorise the amount smoked, and use information on the time from waking until smoking the first cigarette. This latter variable has a high level of missingness. Together these categorical variables allow calculation of \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3307335/}{the heaviness of smoking index}.  

# Using hseclean

## Load and clean data

The data is stored in `X:/ScHARR/PR_Consumption_TA/Data/`. The following code will read, clean, filter and combine the data. 

```{r read data, eval = F}
# Install and load the package
install.packages("X:/ScHARR/PR_STAPM/Code/R_packages/hseclean_0.1.0.zip", repos = NULL)
library(hseclean)

# Write a bespoke function that does just the cleaning jobs required.
cleandata <- function(data) {
  data <- clean_surveyweights(data)
  data <- clean_age(data)
  data <- clean_demographic(data)
  data <- smk_status(data)
  data <- smk_former(data)
  data <- smk_life_history(data)
  data <- smk_amount(data)
  data <- select_data(
    data, ages = 12:89, years = 2001:2016,
    
    # The variables to retain
    keep_vars = c("wt_int", "psu", "cluster", "year", 
                  "age", "age_cat", "censor_age", "sex", "imd_quintile",
                  "cig_smoker_status", "smk_start_age", "years_since_quit"),
    
    # The variables that must have complete cases
    complete_vars = c("cig_smoker_status", "wt_int", "psu", "cluster", "year", "censor_age")
  )
return(data)
}

hse_data <- combine_years(list(
  cleandata(read_2014()), cleandata(read_2015()), cleandata(read_2016())
))

# clean the survey weights
hse_data <- clean_surveyweights(hse_data)

# change some variable names
setnames(hse_data, c("smk_start_age", "cig_smoker_status", "years_since_quit"),
         c("start_age", "smk.state", "time_since_quit"))

# check that all current smokers have an amount smoked per data that is greater than zero
testthat::expect_equal(nrow(hse_data[smk.state == "current" & cigs_per_day == 0]), 0,
                       info = "some current smokers smoke 0 cigs per day")

# select only data with a coherent time since quit
hse_data[ , time_since_quit := as.double(ceiling(time_since_quit))]
hse_data <- hse_data[!(smk.state == "former" & time_since_quit < 1)]
```

## Summarise data
Taking the survey design into account is important when estimating the mean and confidence intervals around summary statistics computed from the data i.e. it is not possible to accurately estimate sampling error without accounting for survey design. The `survey` R package [@Rsurvey] has a collection of functions that incorporate survey design into the calculation of summary statistics. The `survey` package is used by the function `prop_summary()` in `hseclean` to estimate the uncertainty around proportions calculated from a binary variable - `prop_summary()` was designed to simplify the process of estimating smoking prevalence from the HSE data, stratified by a specified set of variables.   


Using `prop_summary()`, calculate the proportion of smokers, stratified by year, sex and quintiles of the Index of Multiple Deprivation.  

```{r ex2, eval = F}
prop_smokers <- prop_summary(
  data = hse_data,
  var_name = "smk.state",
  levels_1 = "current",
  levels_0 = c("former", "never"),
  strat_vars = c("year", "sex", "imd_quintile")
)

```

```{r smoking trends, eval = T, echo=F, warning=F, out.extra='', fig.width = 7, fig.height = 6, fig.cap = "Trends in the proportion of smokers."}

readRDS("X:/ScHARR/PR_STAPM/Code/R_packages/hseclean/vignettes/current_smoking_sex_imd_trend.rds")

```  

## Impute data

### Imputation during data cleaning
There has already been some filling-in of missing data done by the cleaning functions described above.  

- For 2015+, the `clean_age()` function has randomly assigned single years of age within each age category.  
- If someone is classified as a current smoker, is younger than 16 and has missing data on the time between waking and having their first cigarette, we assume that this time is one hour or more.  
- If someone is younger than age 16 and has missing data on their number of children (for survey years prior to 2015), we assume that they had no children.  
- If someone of any age is classified as a current smoker but has a missing value for the amount smoked, then we fill that missing value with the average amount smoked within a year, age, sex and IMD quintile subgroup.

### Selecting complete cases
In the data preparation code above, we also have the option to filter the data to retain only complete cases for certain variables. Currently, we retain only records with full information on age, sex, smoking status and the survey design variables. We then multiply impute missing values for key socio-demographic variables.  

### Multiple imputation
To conduct our multiple imputation, we use the R package `mice` [@Rmice]. Since the process of running the multiple imputation can take a long time and consume a lot of RAM, it might be easiest to run on the `heta_study` virtual machine. There is a range of `mice` documentation and tutorials online that is worth getting into. In `hseclean`, multiple imputation is implemented by the `impute_data_mice()` function, although this is currently written rather simply and could be checked and developed to include more optionality and potentially also diagnostic checks.  

`mice` basically fits a chained series of regression equations that predict the missing values of successive variables based on the existing relationships between those variables and selected other variables in the data. The current usage only imputes categorical variables, which could be one of three types: "logreg" - binary Logistic regression; "polr" - ordered Proportional odds model; "polyreg" - unordered Polytomous logistic regression.  

In running the multiple imputation, the number of desired iterations of the imputed data is selected (I current generate five to keep the size of the resulting imputed data manageable), and the variables to either be predicted or to inform the prediction are selected. If a variable is just going to inform the prediction of the other variables but is not going to be predicted itself, then the model type is set to "", otherwise to one of "logreg", "polr" or "polyreg".  

```{r mi example, eval = F}
# First inspect variables
table(hse_data$kids, useNA = "ifany")

# Set a broader age category variable
hse_data[ , agegroup := 
            c("12-16", "16-17", "18-24", "25-34", "35-49", "50-64", "65-74", "75-89")[
              findInterval(age, c(-10, 16, 18, 25, 35, 50, 65, 75, 1000))]]

# Run the imputation (takes a long time)
imp <- impute_data_mice(data = hse_data,
                        var_names = c("smk.state", "agegroup", "sex", 
                                      "imd_quintile", "degree", "kids", "income5cat",
                                      "relationship_status", "employ2cat", "social_grade"),
                        var_methods = c("", "", "", 
                                        "polr", "logreg", "polr", "polr",
                                        "polyreg", "logreg", "logreg"),
                        n_imputations = 5)

# imp$data is a single data.table containing all 5 imputed versions of the data
hse_data_smoking <- copy(imp$data)
```

# References
















































































































